### 2025-05-14 docs/article/Sections/benchmark.tex
small improvements and fixes to chapters 4 and 5 of the article; changed journal to Software.

The benchmark is designed with a modular architecture, separating the
\textit{view} and \textit{model} layers from the \textit{controller} layer
\citeauthor{Bucanek2009}, which allows for easy extension and integration of new
template engines and frameworks. It also includes a set of tests to ensure the
correctness of implementations and to validate the \textit{HTML} output.

The benchmark includes two different data models, defined as follows:

@@

The Spring WebFlux implementation uses Project Reactor to support a reactive
programming model: each method returns a \texttt{Flux<String>} as the response
body, which acts as a publisher that progressively streams the HTML content to
the client. It also includes methods using Kotlin coroutines and other
asynchronous mechanisms supported by template engines, such as the
\texttt{writeAsync} method from HtmlFlow, which enables non-blocking I/O
using \textit{continuations}. The aforementioned blocking template engines are used in
the context of Virtual Threads or alternative coroutine dispatchers, allowing
the handler thread to be released and reused for other requests.

The Spring MVC implementation uses handlers based solely on the blocking
interface of the \texttt{Observable} class. To enable PSSR in this context, we utilize
the \texttt{StreamingResponseBody} interface, which allows the application to
write directly to the response \texttt{OutputStream} without blocking the servlet
container thread. According to the Spring documentation, this class is \textit{a
controller method return value type for asynchronous request processing where
the application can write directly to the response OutputStream without holding
up the Servlet container thread} \citeauthor{Spring-StreamingResponseBody}.

In Spring MVC, \texttt{StreamingResponseBody} enables asynchronous writing relative to
the request-handling thread, but the underlying I/O remains
blocking—specifically the writes to the \texttt{OutputStream}. When using Virtual
Threads, the I/O operations are more efficient, as they are executed in the
context of a lightweight thread. Most of the computation is done in a separate
thread from the one that receives each request; we use a thread pool
\texttt{TaskExecutor} to process requests, allowing the application to scale and
handle multiple clients more efficiently as opposed to the default
\texttt{TaskExecutor} implementation, which tries to create a thread for each request.

However, the Spring MVC implementation does not effectively support PSSR for
these templates, as HTML content is not streamed progressively to the client.
This is because the response is only sent once the content written to the
\texttt{OutputStream} exceeds the output buffer size, which defaults to 8KB\@. As a
result, the client receives the response only after the entire HTML content is
rendered, defeating the purpose of PSSR in this context. We attempted to reduce
the buffer size by configuring it via an \texttt{HttpFilter}, but this did not affect
the observed behavior.

The Quarkus implementation also uses handlers based on the blocking interface
of the \texttt{Observable} class. It implements the \textbf{StreamingOutput} interface
from the JAX-RS specification to enable PSSR, allowing HTML content to be
streamed to the client. While \texttt{StreamingOutput} also uses blocking I/O, it
operates on Vert.x worker threads, which prevents blocking of the event loop.
When Virtual Threads are used, the I/O operations are handled efficiently, as
they are executed in lightweight threads.

>>>>

The benchmark is designed with a modular architecture, separating the
\textit{view} and \textit{model} layers from the \textit{controller} layer
\cite{Bucanek2009}, which allows for easy extension and integration of new
template engines and frameworks. It also includes a set of tests to ensure the
correctness of implementations and to validate the \textit{HTML} output.

The benchmark includes two different data models, defined as \texttt{Presentation} and 
\texttt{Stock} shown in \autoref{lst:data-models}.

@@

The aforementioned blocking template engines are used in
the context of Virtual Threads or alternative coroutine dispatchers, allowing
the handler thread to be released and reused for other requests.

The Spring WebFlux core implementation uses Project Reactor to support a reactive
programming model: each method returns a \texttt{Flux<String>} as the response
body, which acts as a publisher that progressively streams the HTML content to
the client. It also includes methods using Kotlin coroutines and other
asynchronous mechanisms supported by template engines, such as the
\texttt{writeAsync} method from HtmlFlow, which enables non-blocking I/O
using \textit{continuations}. 
The Spring MVC implementation uses handlers based solely on the blocking
interface of the \texttt{Observable} class. To enable PSSR in this context, we utilize
the \texttt{StreamingResponseBody} interface, which allows the application to
write directly to the response \texttt{OutputStream} without blocking the servlet
container thread. According to the Spring documentation, this class is \textit{a
controller method return value type for asynchronous request processing where
the application can write directly to the response OutputStream without holding
up the Servlet container thread} \cite{Spring-StreamingResponseBody}.

In Spring MVC, \texttt{StreamingResponseBody} enables asynchronous writing relative to
the request-handling thread, but the underlying I/O remains
blocking—specifically the writes to the \texttt{OutputStream}. When using Virtual
Threads, the I/O operations are more efficient when compared to platform threads,
as they are executed in the context of a lightweight thread. Most of the computation is done in a separate
thread from the one that receives each request; we use a thread pool
\texttt{TaskExecutor} to process requests, allowing the application to scale and
handle multiple clients more efficiently as opposed to the default
\texttt{TaskExecutor} implementation, which tries to create a thread for each request.

However, the Spring MVC implementation does not effectively support PSSR for
these templates, as HTML content is not streamed progressively to the client.
This is because the response is only sent once the content written to the
\texttt{OutputStream} exceeds the output buffer size, which defaults to 8KB\@. As a
result, the client receives the response only after the entire HTML content is
rendered, defeating the purpose of PSSR in this context.

The Quarkus implementation also uses handlers based on the blocking interface
of the \texttt{Observable} class. It implements the \texttt{StreamingOutput} interface
from the JAX-RS specification to enable PSSR, allowing HTML content to be
streamed to the client. While \texttt{StreamingOutput} also uses blocking I/O, it
operates on Vert.x worker threads, which prevents blocking of the event loop.
When Virtual Threads are used, the I/O operations are handled efficiently, as
they are executed in lightweight threads.

### 2025-05-14 docs/article/Sections/results.tex
small improvements and fixes to chapters 4 and 5 of the article; changed journal to Software.

The results in Figure~\ref{fig:presentations-webflux-jmeter} depict the
throughput (number of requests per second) for each template engine, with
concurrent users ranging from 1 to 128, from left to right. The benchmarks
include HtmlFlow using suspendable web templates (HtmlFlow-Susp), Jstachio
using Virtual Threads (Jstachio-Virtual), and Thymeleaf using the reactive View
Resolver driver (Thymeleaf-Rx). \textit{Sync} and \textit{Virtual} represent
the average throughput of the blocking approaches (i.e., KotlinX, Rocker,
Jstachio, Pebble, Freemarker, Trimou, HtmlFlow, and Thymeleaf) when run in the
context of a separate coroutine dispatcher or Virtual Threads, respectively.

We show the HtmlFlow-Susp, Jstachio-Virtual, and Thymeleaf-Rx engines
separately to observe the performance of the non-blocking engines when using
the Reactive, Virtual Thread, and Suspending approaches. The \textit{Sync} and
\textit{Virtual} are aggregated due to the similar performance of different
engines when using those approaches.

>>>>

The results in Figure~\ref{fig:presentations-webflux-jmeter} depict the
throughput (number of requests per second) for each template engine, with
concurrent users ranging from 1 to 128, from left to right. The benchmarks
include HtmlFlow using suspendable web templates (HtmlFlow-Susp), Jstachio
using Virtual Threads with the Iterable interface (Jstachio-Virtual), and Thymeleaf using the reactive View
Resolver driver (Thymeleaf-Rx). \textit{Sync} and \textit{Virtual} represent
the average throughput of the blocking approaches (i.e., KotlinX, Rocker,
Jstachio, Pebble, Freemarker, Trimou, HtmlFlow, and Thymeleaf) when run in the
context of a separate coroutine dispatcher or Virtual Threads, respectively.

We show the HtmlFlow-Susp, Jstachio-Virtual, and Thymeleaf-Rx engines
separately to observe the performance of the non-blocking engines when using
the Suspending, Virtual Thread, and Reactive approaches. The \textit{Sync} and
\textit{Virtual} are aggregated due to the similar performance of different
engines when using those approaches.
